---
title: "Introduction to R"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

There are four basic data types in R (that you need to know): character, numeric, integer, logical.
There are four basic ways to store data in R: vectors, data frames, matrices, and lists.

You can learn about these data types below    

#### Numerics and integers  

```{r}
#to store data in some variable name, use either = or <-
#to save the number 5 as "number":
number <- 5

#print number:
print(number)  

#find out the class of number:
class(number)

#change number to an integer and re-save it as number2:
number2 <- as.integer(number)

#inspect the class of number2 to see that it is an integer:
class(number2)
```
    
#### Characters  

```{r}
#save a character string in a variable named message
message <- "welcome"

#print message:
print(message)

#inspect the class of message:
class(message)
```

  
#### Logicals   
Logical data is either TRUE or FALSE  
In R, TRUE=1 and FALSE=0   
```{r}
#save the logical TRUE as a variable called outcome:  
outcome <- TRUE  

#print outcome
print(outcome)

#inspect class of outcome
class(outcome)

#note that, weirdly, outcome+outcome=2  
outcome+outcome
```
  
In R, you can test a statement to see if it is TRUE or FALSE. Note that R allows you to make comparisons accross variable types: integers may be compared to numerics and logicals may be compared to integers/numerics. For characters, comparatives are assessed using alphabetical order (letters earlier in the alphabet are "smaller"):    
  1) == means "is equal to"  
  2) != means "is not equal to"  
  3) > means "greater than"; >= means "greater than or equal to"  
  4) < means "less than; <= means "less than or equal to"  
  
```{r}
#is 5 equal to 3?
5==3

#is 5 less than 3?
5<3

#is 5 greater than 3?
5>3

#is 5 greater than 5?
5>5

#is 5 greater than or equal to 5?
5>=5

#is 5 equal to 5?
5==5

#is "hello" equal to "hello"?
"hello" == "hello"

#is "hello" equal to "goodbye"?
"hello" == "goodbye"

#is "hello" greater than "goodbye"? (in other words is "hello" after "goodbye" alphabetically?)
"hello">"goodbye"  

#is TRUE == 1?
TRUE==1

#is FALSE==0?
FALSE==0
```
  
#### Vectors  
```{r}
# The easiest way to create a vector is by using the c() function
vec1 <- c(2,3,4,5)
print(vec1)

# Note that, if you include multiple data types in a vector, R will change all values to the same data type
vec2 <- c("hello",7,3)
print(vec2)

# For values in a row, we can also use a colon:
vec3 <- 2:5
print(vec3)

# We can use the c() function to combine pre-saved vectors:
vec4 <- c(vec1,vec3)
print(vec4)

# Other shortcuts for creating vectors in R   

#use the rep() function to repeat values. Inspect the following to see how it works!
rep(x=0,times=5) #create a vector of 5 zeros
rep(x=c(1,2,3),times=5) #create a vector with five repeats of 1,2,3
rep(c(1,2,3),each=2,times=5) #every value in 1,2,3 is repeated twice, and then that sequence is repeated 5 times

```

