---
title: "Quantitative_Methods_Assignment"
author: "Firstname Lastname"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Two-sample t-tests 

If we want to compare two independent sample means for data that is not paired (for example, if we had prices for 100 drinks at Starbucks and 100 drinks at Dunkin Donuts, but they were not paired by drink type), we can use an independent sample T-test.   

This is done in much the same way as before, except now we need to calculate $\bar{x_1}$ (the mean of group 1), $\bar{x_2}$ (the mean of group 2), calculate their difference, and then construct a T-test comparing this value to a null hypothesis that $\bar{x_1}-\bar{x_2}=0$. Also, we need a different way of estimating the standard error and need to know how many degrees of freedom to use for our t distribution (see below).    
   
Standard error:  
Using what we know about the properties of variance, we can derive that:  
$$SE^2(\bar{x_1}-\bar{x_2})=SE^2(\bar{x_1})+(-1)^2SE^2(\bar{x_2})= SE^2(\bar{x_1})+SE^2(\bar{x_2})=\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}$$  
So, $SE(\bar{x_1}-\bar{x_2})$ can be estimated using: $$SE(\bar{x_1}-\bar{x_2})=\sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}}$$    

Alternatively, if we think that the two groups have equal variances, we can used the pooled standard deviation formula to get a better estimate of $SE(\bar{x_1}-\bar{x_2})$. The formula is:  
$$s^2_{pooled}=\frac{s_1^2(n_1-1)+s_2^2(n_2-1)}{n_1+n_2-2}$$     


Degrees of freedom:  
In general, we use the smaller of $n_1-1$ and $n_2-1$ as the degrees of freedom for the t distribution (or if assuming equal variances and using the $s^2_{pooled}$, we can use $n_1+n_2-2$ degrees of freedom).   

```{r}
height <- read.csv("height_sex.csv")

# Average height of men
mean(height$Height[height$Male==1])

# Average height of women
mean(height$Height[height$Male==0])
```

```{r}
# Average height of men
mean_hm <- mean(height$Height[height$Male==1])
var_hm <- var(height$Height[height$Male==1])
N_hm <- sum(height$Male)

# Females
# Average height of women
mean_hw <- mean(height$Height[height$Male==0])
var_hw <- var(height$Height[height$Male==0])
N_hw <- length(height$Male) - sum(height$Male)

sd_comb <- sqrt((var_hm/N_hm)+(var_hw/N_hw))

t <- (mean_hm - mean_hw)/sd_comb
1-pt(t, df = (N_hm+N_hw-2))
```


### Data Exploration
```{r}
summary(lm(Height~as.factor(Male), data=height))
# walk through lm() in this exercise 
```



