---
title: "2352-Statistical Computing Assignment: Simulating the Central Limit Theorem"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We can sample from a variety of distributions in R. To sample from a normal distribution we would use **rnorm**. rnorm() takes in parameters of the distribution: n (the number of observations), mean (the mean of the sample), and sd (the standard deviation) to randomly draw from the distribution. The example below creates a sample of size = 500 with mean = 0 and standard deviation = 1. 

```{r}
# Set seed
set.seed(12345) 

# A sample of 100 observations from a normal distribution with mean = 0 and sd = 1. 
dat1 <- rnorm(200, 0, 1)

# We can plot 
par(mfrow=c(1,2)) #This creates matrix of m rows by n columns to be able to plot more than one plot
plot(density(dat1), main="Density Plot of random sample", cex.main = 0.8)
hist(dat1, main = "Histogram of random sample", ylab="", xlab="", cex.main = 0.8)
```

```{r}
# Draw 20 observations and take the mean of the draw 
draw1 <- sample(dat1, 20)
mean(draw1)
hist(draw1)
```



### The Central Limit Theorem
The **central limit theorem** states that given a population with mean of $\mu$ and a standard deivation of $\sigma$, the sampling distribution of the mean has a mean of $\mu$ and a standard deivation of $\frac{\sigma}{\sqrt{N}}$ and approaces a normal distribution as the sample size on which it is based, N, approaches infinity. 

1. The mean of the sampling distribution of the mean



*** Describe what a uniform distribution is!!!!****
For a unifrom distribution, the probability density function, 

1. Sample from a uniform distribution (use runif) and store your sample. 
```{r}
testing1 <- runif(500, 0, 1)
```

2. Create a sampling distribution of the mean with 20, 100, 1000, and 4000 thousand draws 
