---
title: "2352-Statistical Computing Assignment"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Simulating the Central Limit Theorem

We can sample from a variety of distributions in R. To sample from a normal distribution we would use **rnorm**. rnorm() takes in parameters of the distribution: n (the number of observations), mean (the mean of the sample), and sd (the standard deviation) to randomly draw from the distribution. The example below creates a sample of size = 200 with mean = 0 and standard deviation = 1. 

```{r}
# Set seed allows for reproducible results and for the document to produce the exact sample numbers and samples 
set.seed(12345)  

# A sample of 200 observations from a normal distribution with mean = 0 and sd = 1. 
dat1 <- rnorm(200, 0, 1)

# We can visualize our data using the R functions plot(), density(), and hist()
par(mfrow=c(1,2)) #This creates matrix of m rows by n columns to be able to plot more than one plot
plot(density(dat1), main="Density Plot of random sample", cex.main = 0.8)
hist(dat1, main = "Histogram of random sample", ylab="", xlab="", cex.main = 0.8)
```

```{r}
# Draw 20 observations and take the mean of the draw 
N <- 20
draw1 <- sample(dat1, N)
mean(draw1); var(draw1)/sqrt(1)
hist(draw1)
```

Now draw from 
```{r}
draw_mean <- vector()
for (i in 1:N){
  draw_mean[i] <- mean(sample(dat1, 20))
}
hist(draw_mean)
mean(draw_mean)
```


### The Central Limit Theorem
The **central limit theorem** states that given a population with mean of $\mu$ and a standard deivation of $\sigma$, the sampling distribution of the mean has a mean of $\mu$ and a standard deivation of $\frac{\sigma}{\sqrt{N}}$ and approaces a normal distribution as the sample size on which it is based, N, approaches infinity. 

1. The mean of the sampling distribution of the mean
```{r}

```



*** Describe what a uniform distribution is!!!!****
For a unifrom distribution, the probability density function, 

1. Sample from a uniform distribution (use runif) and store your sample. 
```{r}
testing1 <- runif(500, 0, 1)
plot(density(testing1))
```

2. Create a sampling distribution of the mean with 20, 100, 1000, and 4000 thousand draws 
