---
title: "2352-Statistical Computing Assignment"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Simulating the Central Limit Theorem

### The Central Limit Theorem
The **central limit theorem** states that given a population with mean of $\mu$ and a standard deivation of $\sigma$, the sampling distribution of the mean has a mean of $\mu$ and a standard deivation of $\frac{\sigma}{\sqrt{N}}$ and approaces a normal distribution as the sample size on which it is based, N, approaches infinity. 

1. The mean of the sampling distribution of the mean


We can sample from a variety of distributions in R. To sample from a normal distribution we would use **rnorm**. rnorm() takes in parameters of the distribution: n (the number of observations), mean (the mean of the sample), and sd (the standard deviation) to randomly draw from the distribution. The example below creates a sample of size = 200 with mean = 0 and standard deviation = 1. 

```{r, tidy = TRUE}
# Set seed allows for reproducible results and for the document to produce the exact sample numbers and samples 
set.seed(12345)  

# A sample of 200 observations from a normal distribution with mean = 0 and sd = 1. 
dat1 <- rnorm(200, 0, 1)

# We can visualize our data using the R functions plot(), density(), and hist()
par(mfrow=c(1,2)) #This creates matrix of m rows by n columns to be able to plot more than one plot
plot(density(dat1), main="Density Plot of random sample", cex.main = 0.8) # Creates a density plot of the data
hist(dat1, main = "Histogram of random sample", ylab="", xlab="", cex.main = 0.8) # Creates a histogram of the data 
```

```{r, tidy = TRUE}
# Draw 20 observations and take the mean of the draw 

N <- 20 # Storing the number of observations
draw1 <- sample(dat1, N) # Randomly drawing from the above simulated data N number of times and storing it in draw1
mean(draw1) # Computes the mean of the draw
var(draw1)/sqrt(1) #

par(mfrow=c(1,2))
hist(draw1) # Plots a histogram of the draw
plot(density(draw1)) # Creates a density plot of the draw
```

Now let's create a sampling distribution of the mean of size 50. 
```{r, tidy = TRUE}
N <- 20 # Set the size of sample
ITER <- 50 # Set the number of iterations

# Create empty vector of length ITER to store results 
draw_mean1 <- rep(NA, length=ITER)

# Creates a loop to draw 20 observations from the original sample (dat1), takes the mean of the 20 observations, and stores the mean and iterates over the loop ITER times (in this case 50 times)
for (i in 1:ITER){
  draw_mean1[i] <- mean(sample(dat1, N))
}

mean(draw_mean1)

par(mfrow=c(1,2))
hist(draw_mean1, main="Histogram of sample means (N = 50)", cex.main = 0.8) # Plots a histogram of the sample means
plot(density(draw_mean1), main = "Density plot of sample means (N = 50)", cex.main = 0.8) # Creates a density plot of the sample means 
```

Next we can increase the number of iterations to 100, 1000, and 4000 and see what happens to the sampling distribution as we increase the number of iterations. 
```{r, tidy = TRUE}
N <- 20 # Set the size of sample
ITER <- 100 # Set the number of iterations

# Create empty vector of length ITER to store results 
draw_mean2 <- rep(NA, length=ITER)

# Creates a loop to draw 20 observations from the original sample (dat1), takes the mean of the 20 observations, and stores the mean and iterates over the loop ITER times (in this case 100 times)
for (i in 1:ITER){
  draw_mean2[i] <- mean(sample(dat1, N))
}

mean(draw_mean2)

par(mfrow=c(1,2))
hist(draw_mean2, main="Histogram of sample means (N = 100)", cex.main = 0.8) # Plots a histogram of the sample means
plot(density(draw_mean2), main="Density plot of sample means (N = 100)", cex.main = 0.8) # Creates a density plot of the sample means 
```

```{r, tidy = TRUE}
N <- 20 # Set the size of sample
ITER <- 1000 # Set the number of iterations

# Create empty vector of length ITER to store results 
draw_mean3 <- rep(NA, length=ITER)

# Creates a loop to draw 20 observations from the original sample (dat1), takes the mean of the 20 observations, and stores the mean and iterates over the loop ITER times (in this case 1000 times)
for (i in 1:ITER){
  draw_mean3[i] <- mean(sample(dat1, N))
}

mean(draw_mean3)

par(mfrow=c(1,2))
hist(draw_mean3, main="Histogram of sample means (N = 1000)", cex.main = 0.8)# Plots a histogram of the sample means
plot(density(draw_mean3), main="Density plot of sample means (N = 1000)", cex.main = 0.8) # Creates a density plot of the sample means 
```

```{r, tidy = TRUE}
N <- 20 # Set the size of sample
ITER <- 4000 # Set the number of iterations

# Create empty vector of length ITER to store results 
draw_mean4 <- rep(NA, length=ITER)

# Creates a loop to draw 20 observations from the original sample (dat1), takes the mean of the 20 observations, and stores the mean and iterates over the loop ITER times (in this case 4000 times)
for (i in 1:ITER){
  draw_mean4[i] <- mean(sample(dat1, N))
}

mean(draw_mean4)

par(mfrow=c(1,2))
hist(draw_mean4, main="Histogram of sample means (N = 4000)", cex.main = 0.8) # Plots a histogram of the sample means
plot(density(draw_mean4), main="Density plot of sample means (N = 4000)", cex.main = 0.8) # Creates a density plot of the sample means 
```

```{r}
par(mfrow=c(2,2))
plot(density(draw_mean1), main="Density plot of sample means (N = 50)", cex.main = 0.8)
plot(density(draw_mean2), main="Density plot of sample means (N = 100)", cex.main = 0.8)
plot(density(draw_mean3), main="Density plot of sample means (N = 1000)", cex.main = 0.8)
plot(density(draw_mean4), main="Density plot of sample means (N = 4000)", cex.main = 0.8)
```

*** Describe what a uniform distribution is!!!!****
For a unifrom distribution, the probability density function, 

1. Sample from a uniform distribution (use runif) and store your sample. 
```{r, tidy = TRUE}
testing1 <- runif(500, 0, 1)
plot(density(testing1))
```

2. Create a sampling distribution of the mean with 20, 100, 1000, and 4000 thousand draws 
