---
title: "R Code to Complement Math Review Packet"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Properties of Logarithms

## Practice Problems.
1. Solve the following:
a. $log_e(e^x) = x$
```{r Logs 1a}
# If x = 3
log(exp(3))
```

b. $log_{10}(100) = 2$
```{r Logs 1b}
log10(100)
```

c. $log_{10}(\frac{1}{10}) = -1$
```{r Logs 1c}
log10(1/10)
```

d. $log_{10}(0) =$ No solution
```{r Logs 1d}
log10(0)
```


3c. $log_{10}(5) + log_{10}(2) = 1$
```{r Log 3c}
log10(5) + log10(2)
```

# Matrix Algebra

Type ?matrix in the **Condole** or matrix in **Help** to look at the inputs of the matrix function.

If we want to store the matrix we need to call it "A", for example and store the matrix. 

For example $\textbf{Z} = \begin{bmatrix} 1&2&3\\ 4&5&6\\ 7&8&9\\ \end{bmatrix}$ $\textbf{Y} = \begin{bmatrix} 1&4&7\\ 2&5&8\\ 3&6&9\\ \end{bmatrix}$

```{r Matrix Example}
#Sorting by row
Z <- matrix(data = c(1,2,3,4,5,6,7,8,9), 
             nrow = 3, ncol = 3,  
             byrow = TRUE)

print(Z)

#Sorting by column
Y <- matrix(data = (1:9), 
             nrow = 3, ncol = 3,
             byrow = FALSE)

print(Y)
```

## Practice Problems
1. Solve the following:
  a. $\begin{bmatrix} 2&4&2\\ 1&3&0\\1&6&2\\ \end{bmatrix} +
      \begin{bmatrix} 1&5&0\\ -2&-3&0\\ 1&9&5\\ \end{bmatrix} = 
      \begin{bmatrix} 3&9&2\\ -1&0&0\\ 2&15&7\\ \end{bmatrix}$  
    
```{r Matrix 1a}
M1 <- matrix(data = c(2,4,2,1,3,0,1,6,2),
             nrow = 3, ncol = 3, byrow = TRUE)

M2 <- matrix(data = c(1,5,0,-2,-3,0,1,9,5),
             nrow = 3, ncol = 3, byrow = TRUE)

M1 + M2
```

  b. $\begin{bmatrix} 2&1\\ -2&2\\ 4&2\\ \end{bmatrix}
      \begin{bmatrix} 5&2&-1\\ 3&4&2\\ \end{bmatrix} =
      \begin{bmatrix} 13&8&0\\ -4&4&6\\ 26&16&0\\ \end{bmatrix}$  
      
```{r Matrix 1b}
M3 <- matrix(data = c(2,-2,4,1,2,2), 
             nrow = 3, ncol = 2,
             byrow = FALSE)

M4 <- matrix(data = c(5,3,2,4,-1,2), 
             nrow = 2, ncol = 3,
             byrow = FALSE)

M3 %*% M4
```

c. Let $\textbf{A} =\begin{bmatrix} 1&2\\ 3&5\\ 4&0\\ \end{bmatrix} \mbox{and } 
       \textbf{B} =\begin{bmatrix} 4&4\\ 1&2\\ 7&0\\ \end{bmatrix}  
       \textbf{A}^T \textbf{B} = \begin{bmatrix} 35&10\\ 13&18\\ \end{bmatrix}$ 
       
```{r Matrix 1c}
A <- matrix(data = c(1,2,3,5,4,0),
            nrow = 3, ncol = 2, byrow = TRUE)

B <- matrix(data = c(4,1,7,4,2,0),
            nrow = 3, ncol=2, byrow = FALSE)

# Finding A transpose
tA <- t(A)
print(tA)

tA %*% B
            
```
d. Using the same matrices as in part c, $\textbf{B}^T\textbf{A} = 
\begin{bmatrix} 35&13\\ 10&18\\ \end{bmatrix}$

```{r 1d}
tB <- t(B)

tB %*% A
```

Show that **A** and **B** are inverses:
$\begin{bmatrix} 1&2&1\\ 2&2&0\\ 1&1&1\\ \end{bmatrix}  
  \begin{bmatrix} -1&0.5&1\\ 1&0&-1\\ 0&-0.5&1\\ \end{bmatrix}=
  \begin{bmatrix} 1&0&0\\ 0&1&0\\ 0&0&1\\ \end{bmatrix}$
       
       
```{r Matrix 2}
A <- matrix(data = c(1,2,1,2,2,0,1,1,1),
            nrow = 3, ncol = 3, byrow = TRUE)

#Use the solve() to find the inverse of a matrix
solve(A)

B <- matrix(data = c(-1, 0.5, 1, 1, 0, -1, 0, -0.5, 1),
             nrow = 3, ncol = 3, byrow = TRUE)

solve(B)

A %*% B
```

3. Suppose that $A$ is a 4x3 matrix and $B$ is a $3x8$ matrix.  
```{r Matrix 3}
A <- matrix (data= (1:12), nrow = 4, ncol = 3, byrow = FALSE)
B <- matrix (data = (1:24), nrow = 3, ncol = 8, byrow = TRUE)
```
  a. $\textbf{AB}$ exists and is 4x8 matrix. 
```{r Matrix 3a}
A %*% B

dim(A%*%B)
```
  b. $\textbf{BA}$ does not exist. 
```{r}
# The following produce errors
# B %*% A
# dim(B %*% A)
```

4. The determinant of $det\begin{bmatrix} 1&-2\\ 4&3\\ \end {bmatrix} = 11$
```{r Matrix 4}
M5 <- matrix(data = c(1, -2, 4, 3), nrow = 2, ncol = 2, byrow = TRUE)

det(M5)
```


      