---
title: "R Code to Complement Math Review Packet"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Properties of Logarithms

## Practice Problems.
1. Solve the following:
a. $log_e(e^x) = x$
```{r}
# If x = 3

log(exp(3))
```

b. $log_{10}(100) = 2$
```{r}
log10(100)
```

c. b. $log_{10}(\frac{1}{10}) = -1$
```{r}
log10(1/10)
```

d. b. $log_{10}(0) = \mbox{ No solution} $

3c. $log_{10}(5) + log_{10}(2) = 1$
```{r}
log10(5) + log10(2)
```

# Matrix Algebra

Type ?matrix in the **Condole** or matrix in **Help** to look at the inputs of the matrix function.

If we want to store the matrix we need to call it "A", for example and store the matrix. 

For example $\textbf{Z} = \begin{bmatrix} 1&2&3\\ 4&5&6\\ 7&8&9\\ \end{bmatrix}$ $\textbf{Y} = \begin{bmatrix} 1&4&7\\ 2&5&8\\ 3&6&9\\ \end{bmatrix}$

```{r}
#Sorting by row
Z <- matrix(data = c(1,2,3,4,5,6,7,8,9), 
             nrow = 3, ncol = 3,  
             byrow = TRUE)

print(Z)

#Sorting by column
Y <- matrix(data = (1:9), 
             nrow = 3, ncol = 3,
             byrow = FALSE)

print(Y)
```

## Practice Problems
1. Solve the following:
  a. $\begin{bmatrix} 2&4&2\\ 1&3&0\\1&6&2\\ \end{bmatrix} +
      \begin{bmatrix} 1&5&0\\ -2&-3&0\\ 1&9&5\\ \end{bmatrix} = 
      \begin{bmatrix} 3&9&2\\ -1&0&0\\ 2&15&7\\ \end{bmatrix}$  
    
```{r}
M1 <- matrix(data = c(2,4,2,1,3,0,1,6,2),
             nrow = 3, ncol = 3, byrow = TRUE)

M2 <- matrix(data = c(1,5,0,-2,-3,0,1,9,5),
             nrow = 3, ncol = 3, byrow = TRUE)

M1 + M2
```

  b. $\begin{bmatrix} 2&1\\ -2&2\\ 4&2\\ \end{bmatrix}
      \begin{bmatrix} 5&2&-1\\ 3&4&2\\ \end{bmatrix} =
      \begin{bmatrix} 13&8&0\\ -4&4&6\\ 26&16&0\\ \end{bmatrix}$  
      
```{r}
M3 <- matrix(data = c(2,-2,4,1,2,2), 
             nrow = 3, ncol = 2,
             byrow = FALSE)

M4 <- matrix(data = c(5,3,2,4,-1,2), 
             nrow = 2, ncol = 3,
             byrow = FALSE)

M3 %*% M4
```

c. Let $\textbf{A} =\begin{bmatrix} 1&2\\ 3&5\\ 4&0\\ \end{bmatrix} and 
       \textbf{B} =\begin{bmatrix} 4&4\\ 1&2\\ 7&0\\ \end{bmatrix}$ Calculate $A^TB$         
      
      